---
title: "Lab11: Genome informatics and high throughput sequencing"
author: "Duy An Le (PID:A16400411)"
format: pdf
---

## Read in file:

```{r}
geneexp <- read.table("geneexpression.txt")
head(geneexp)
```
>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
library(dplyr)
aa = geneexp %>% filter(geno=="A/A")
ag = geneexp %>% filter(geno=="A/G")
gg = geneexp %>% filter(geno=="G/G")

aaplot <- boxplot(aa$exp, plot=F)
agplot <- boxplot(ag$exp, plot=F)
ggplot <- boxplot(gg$exp, plot=F)
```

The sample size of A/A genotype is `r nrow(aa)` and the median expression level is `r aaplot$stats[3]`. The sample size of the A/G genotype is `r nrow(ag)` and the median expression level is `r agplot$stats[3]`. The sample size of the G/G genotype is `r nrow(gg)` and the median expression level is `r ggplot$stats[3]`.


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

Make a boxplot with ggplot
```{r}
ggplot(geneexp) + aes(geno, exp, fill=geno) + geom_boxplot(notch=T)
```

It seems that the relative expression level of ORMDL3 in A/A genotypes is higher than that of the G/G genotypes. Having the SNP seems to affect the expression of ORMDL3 and may be correlated with having asthma. 
